Time invariant analysis of epidemics with \pkg{EpiCompare}

### 2021-01-15 jss/EpiCompare-paper/EpiCompare-paper.tex
mainly citations, cleaning up references, and added proof.

\usepackage{amsmath}

@@

By definition, \(R_0\) is the number of expected secondary infections
when a primary infection is introduced to a susceptible population.
\(R_0\) is also, maybe, the most famous \textit{time-invariant}
numerical summary of an epidemic, which allows epidemics to be compared
to one another in both different time and geographic scales. For
example, \(R_0\) for Covid-19 is estimated to be between 2-3, seasonal
influenza between 1.2-2, and modern measles outbreaks between 5-6
\citep{midas2020,biggerstaff2014,namee2018}.

@@

With regards to traditional epidemic \(state\) vs.~\(time\) plots,
\(R_0\) is difficult to visualize, especially with respect from one
epidemic to another. For example, consider the scenarios where the first
epidemic is generated from a SIR model with \((S(0) = 990, I(0) = 10)\),
\(\beta_1 = 0.3\) and \(\gamma_1 = 0.15\), and the second epidemic is
generated from a SIR model with \((S(0) = 990, I(0) = 10)\),
\(\beta_2 = 0.24\) and \(\gamma_1 = 0.12\) over 40 days. Both epidemics
have the same value of
\(R_0 = \beta_1/ \gamma_1 = \beta_2 / \gamma_2 = 2\). The epidemic
trajectories are shown in the \(state\) vs.~time plots in Figure
\ref{fig:sir-ex}. At a glance, we may assume that Model 1 has a larger
\(R_0\) than Model 2 because the peak of infection occurs more quickly
than in Model 2. On the other hand, we may think Model 2 has a larger
\(R_0\) because it has a slightly larger peak of infection than Model 1.

@@

However, when we plot the trajectories as a single curve using the
ternary plot in the time-invariant view, we immediately see a different
story. In this time-invariant view in Fig. \ref{fig:ternary-ex}, the
points seem to overlap and form the same trajectory. In other words, we
can see there is something fundamentally linking these two different
epidemics, and this fundamental link turns out to be \(R_0\).

More formally, let two Kermack and McKendrick (see Eq.
\eqref{eq:sir-ode}) SIR models be denoted \((S_1(t), I_1(t), R_1(t))\)
and \((S_2(t), I_2(t), R_2(t))\), respectively, for \(t > 0\). Assume
both models have initial values \((S(0), I(0), R(0))\). Let
\(R_0 = \frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}\) where
\(\beta_i\) and \(\gamma_i\) are the average infection rate and recovery
rate, respectively, for SIR model \(i=1, 2\). Equivalently,
\(\beta_2 = a \beta_1\) if and only if \(\gamma_2 = a \gamma_1\) for
some \(a > 0\).

@@

The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent
result from \cite{Harko2014} and is shown in detail in Appendix
\ref{app:proof}. The consequence of Theorem \ref{thm:sir-scale} is that
for two SIR models that have the same initial percent of individuals in
each state and \(R_0\) then for every point on the epidemic path of the
first SIR model is also a point on the epidemic path of the second SIR
model. Taking the sample simulations from Fig.
\ref{fig:different-scales-standard}, Fig.
\ref{fig:different-scales-tern} presents these two models in a ternary
plot. This means with our time-invariant ternary plot, that at a glance,
we can tell if two epidemics have different values of \(R_0\).

@@

But time-invariant analysis is also useful for epidemic models with more
than three states. The constraints in three dimensions that are met with
the SIR model (that is
\(\sum_{i=1}^3 (\text{number in state}(i)) = N(t)\)) actually represents
a space of 3d simplices, and the ternary plot specifically represents
these simplices, after scaling to examine such values as proportions
between 0 and 1 (ternary plots are known as a 3d unit simplex due to
it's scaling). This same scaling for larger models (i.e.~with more
states) can be done onto different simplexes. In this package we present
tools to help compare models (mostly through simulations). s tool
compares these objects after projecting them into a one-dimension-fewer
space through the simplexical structure of the data.

@@

The first step of most data analysis is cleaning up the data to be
explored. There are multiple ways to collect the epidemiological data.
Sometimes individual records with times of different states of the
epidemic (infection, recovery, etc.) as well as individual information
like network structure, location, and-sub population information will be
collected, whereas other data collections will focus
population/sub-population counts of individuals in each epidemic
state\footnote{\@Shannon include citations}. We focus on understanding
the overall impact on the population/sub-populations we provide a
function to transform information about each agent's start time of each
state (e.g.~start of infection, etc). This transformation between from
agent information to aggregate information is very useful for seeing the
overall trend of the epidemic (and how it impacts different
subpopulations). Our tools aim to allow the user to easily group agents
and define new subpopulations to explore. This is important as many case
studies have highlighted the usefulness to identify differing
subpopulations\footnote{\@Shannon cite} and many state based models
provide for subpopulation based states in their
analysis\footnote{\@Shannon cite}.

@@

With raw data, getting to know our data now-a-days very frequently means
figuring out good combinations of visualizations, numerical summaries
and subsetting. An expert coder can start from
\texttt{agents\_to\_aggregate} to successfully do this in many ways, but
we've also developed tools to rapidly explore your data if your have a 3
unique states models (like the SIR model). Our \texttt{geom\_aggregate}
provides a rapid way to explore different subpopulations experience of
the epidemic, and combines the ideas behind agents\_to\_aggregate for
the SIR case, with \texttt{geom\_path} and \texttt{coord\_term} to
visualize any number of groups epidemic trajectory in 3d simplex space
using \texttt{ggplot2} and \texttt{ggtern}
\citep{Wickham2016, Hamilton2018}. Visualization tools for SIR models
were developed because (1) SIR models are the most common and basic
state-based models\footnote{\@Shannon: cite} and (2) our simplex
representation of these epidemics emphasizes a ``time-invarance''
representation of the data (for a refresher see Section
\ref{sec:time-invariant}).

@@

In this section, we highlight a number of the functionalities available
in \kg{EpiCompare}. These functionalities include data cleaning,
visualization, simulation, and comparison, in accordance with the data
analysis pipeline \ref{fig:pipeline}. We show a full data analysis from
beginning to end that can be accomplished in a streamlined and
standardized manner.

@@

We first try to model the Hagelloch data with an SIR model (see Eq.
\ref{eq:sir}). In our vignette, we show how to fit a stochastic model
via maximum likelihood and simulate from the model with those best fit
parameters. Our function \code{simulate_agents()} generates individual
level data according to discrete multinomial draws, which depend on the
number of individuals in each state at the previous time step and a
matrix of transition probabilities. For example, the below code
generates 100 simulations of an outbreak of a diseease with one initial
infector in a population of \(n= 188\) individuals.

@@

The result of our simulation is the object \code{agents} which is a
18800 \(\times\) 5 which details the time of entry into the \(S\),
\(I\), and \(R\) states for a given simulation. Before we examine the
results of this simple SIR model, we will also examine another, more
sophisticated SIR model, this time from the package \pkg{EpiModel}.
Briefly, this model first fits a contact network to the set of
indivduals, where the class of the student is a covariate. The model
then simulates a SIR-epidemic on that network.

@@

The output of this model is \code{epimodel_sir}, an object of class
netsim, which contains a plethora of modeling information. We provide
the function \code{fortify_aggregate()}, which can take objects from
specialized classes of modeling output and transform it into a
tidy-style data frame.

@@

\hypertarget{helpful-rmd-tricks}{%
\section{Helpful Rmd tricks}\label{helpful-rmd-tricks}}

\hypertarget{rmd-code-formatting-info}{%
\subsection{RMD Code formatting info}\label{rmd-code-formatting-info}}

This is the Section \ref{sec:intro}. This template demonstrates some of
the basic LaTeX that you need to know to create a JSS article.

In general, don't use Markdown, but use the more precise LaTeX commands
instead:

\begin{itemize}
\item
  \proglang{Java}
\item
  \pkg{plyr}
\end{itemize}

One exception is inline code, which can be written inside a pair of
backticks (i.e., using the Markdown syntax).

If you want to use LaTeX commands in headers, you need to provide a
\texttt{short-title} attribute. You can also provide a custom identifier
if necessary. See the header of Section \ref{r-code} for example.

\subsection[R code]{RMD \proglang{R} code}\label{r-code}

Can be inserted in regular R markdown blocks.

hags hags hags \cite{Neal2004}

\begin{CodeChunk}
\begin{CodeInput}
R> x <- 1:10
R> x
\end{CodeInput}
\begin{CodeOutput}
 [1]  1  2  3  4  5  6  7  8  9 10
\end{CodeOutput}
\end{CodeChunk}

>>>>

\usepackage{amsmath} \usepackage{amsthm}

@@

By definition, \(R_0\) is the number of expected secondary infections
when a primary infection is introduced to a susceptible population.
\(R_0\) is also, maybe, the most famous \textit{time-invariant}
numerical summary of an epidemic, which allows epidemics to be compared
to one another in both different time and geographic scales. For
example, \(R_0\) for Covid-19 is estimated to be between 2-3, seasonal
influenza between 1.2-2, and modern measles outbreaks as big as 12
\citep{midas2020,biggerstaff2014,namee2018}.

@@

With regards to traditional epidemic \(state\) vs.~\(time\) plots,
\(R_0\) is difficult to visualize, especially with respect from one
epidemic to another. For example, consider the scenarios where the first
epidemic is generated from a SIR model with \((S(0) = 990, I(0) = 10)\),
\(\beta_1 = 0.3\) and \(\gamma_1 = 0.15\), and the second epidemic is
generated from a SIR model with \((S(0) = 990, I(0) = 10)\),
\(\beta_2 = 0.24\) and \(\gamma_1 = 0.12\) over 40 days. Both epidemics
have the same value of
\(R_0 = \beta_1/ \gamma_1 = \beta_2 / \gamma_2 = 2\). The epidemic
trajectories are shown in the \(state\) vs.~time plots in Figure
\ref{fig:different-scales-standard}. At a glance, we may assume that
Model 1 has a larger \(R_0\) than Model 2 because the peak of infection
occurs more quickly than in Model 2. On the other hand, we may think
Model 2 has a larger \(R_0\) because we may think the number of
infections in that model has not yet peaked at time 15.

@@

However, when we plot the trajectories as a single curve using the
ternary plot in the time-invariant view, we immediately see a different
story. In this time-invariant view in Fig.
\ref{fig:different-scales-tern}, the points seem to overlap and form the
same trajectory. In other words, we can see there is something
fundamentally linking these two different epidemics, and this
fundamental link turns out to be \(R_0\).

More formally, let two Kermack and McKendrick (see Eq.\eqref{eq:sir-ode}
) SIR models be denoted \((S_1(t), I_1(t), R_1(t))\) and
\((S_2(t), I_2(t), R_2(t))\), respectively, for \(t > 0\). Assume both
models have initial values \((S(0), I(0), R(0))\). Let
\(R_0 = \frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}\) where
\(\beta_i\) and \(\gamma_i\) are the average infection rate and recovery
rate, respectively, for SIR model \(i=1, 2\). Equivalently,
\(\beta_2 = a \beta_1\) if and only if \(\gamma_2 = a \gamma_1\) for
some \(a > 0\).

@@

The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent
result from \cite{Harko2014} and is shown in detail in Proof
\ref{proof:thm}. The consequence of Theorem \ref{thm:sir-scale} is that
for two SIR models that have the same initial percent of individuals in
each state and \(R_0\) then for every point on the epidemic path of the
first SIR model is also a point on the epidemic path of the second SIR
model. Taking the sample simulations from Fig.
\ref{fig:different-scales-standard}, Fig.
\ref{fig:different-scales-tern} presents these two models in a ternary
plot. This means with our time-invariant ternary plot, that at a glance,
we can tell if two epidemics have different values of \(R_0\).

@@

But time-invariant analysis is also useful for epidemic models with more
than three states. The constraints in three dimensions that are met with
the SIR model (there are three epidemic states and for each time step,
the sum of total number of individuals in each state is the population
size at that time\$) actually represents a space of 3d simplices, and
the ternary plot specifically represents these simplices, after scaling
to examine such values as proportions between 0 and 1 (ternary plots are
known as a 3d unit simplex due to it's scaling). This same scaling for
larger models (i.e.~with more states) can be done onto different
simplexes. In this package we present tools to help compare models
(mostly through simulations). s tool compares these objects after
projecting them into a one-dimension-fewer space through the simplexical
structure of the data.

@@

The first step of most data analysis is cleaning up the data to be
explored. There are multiple ways to collect the epidemiological data.
Sometimes individual records with times of different states of the
epidemic (infection, recovery, etc.) as well as individual information
like network structure, location, and sub-population information will be
collected, whereas other data collections will focus aggregate counts of
individuals in each epidemic state. In fact, usually only the number of
new infections at each time step is observed and the compartment totals
are then imputed from that along with other information about the
disease and the population of interest. We focus on understanding the
overall impact of an outbreak at the aggregate/population level. We
provide a function to transform information about each
individual/agent's start time of each state (e.g.~start of infection,
etc). This transformation between agent information to aggregate
information is useful for seeing the overall trend of the epidemic (and
how it impacts different subpopulations). Our tools aim to allow the
user to easily group agents and define new subpopulations to explore.
This is important as case studies highlight the usefulness of
identifying differing subpopulations (e.g.~age, sex) and many state
based models provide for subpopulation-based states in their analysis
\citep{rvachev1985,anderson1992,worby2015}.

@@

With raw data, getting to know our data now-a-days very frequently means
figuring out good combinations of visualizations, numerical summaries
and subsetting. An expert coder can start from
\texttt{agents\_to\_aggregate} to successfully do this in many ways, but
we've also developed tools to rapidly explore your data if your have a 3
unique states models (like the SIR model). Our \texttt{geom\_aggregate}
provides a rapid way to explore different subpopulations experience of
the epidemic, and combines the ideas behind agents\_to\_aggregate for
the SIR case, with \texttt{geom\_path} and \texttt{coord\_term} to
visualize any number of groups epidemic trajectory in 3d simplex space
using \texttt{ggplot2} and \texttt{ggtern}
\citep{Wickham2016, Hamilton2018}. Visualization tools for SIR models
were developed because (1) SIR models are some of the most common and
basic epidemic state-based models and (2) our simplex representation of
these epidemics emphasizes a ``time-invarance'' representation of the
data (for a refresher see Section \ref{sec:time-invariant}).

@@

In this section, we highlight a number of the functionalities available
in \pkg{EpiCompare}. These functionalities include data cleaning,
visualization, simulation, and comparison, in accordance with the data
analysis pipeline \ref{fig:pipeline}. We show a full data analysis from
beginning to end that can be accomplished in a streamlined and
standardized manner.

@@

We first try to model the Hagelloch data with a stochastic SIR model,
which we refer to as the `simple SIR.' In our vignette, we show how to
fit this simple SIR model via maximum likelihood and simulate from the
model with those best fit parameters. Our function
\code{simulate_agents()} generates individual level data according to
discrete time multinomial draws, which depend on the number of
individuals in each state at the previous time step and a matrix of
transition probabilities. For example, the below code generates 100
simulations of an outbreak of a disease with one initial infector in a
population of \(n= 188\) individuals.

@@

The result of our simulation is the object \code{agents} which is a
18800 \(\times\) 5 data frame, which details the time of entry into the
\(S\), \(I\), and \(R\) states for a given simulation. Before we examine
the results of this simple SIR model, we will also examine another, more
sophisticated SIR model, this time from the package \pkg{EpiModel}.
Briefly, this model first fits a contact network to the set of
indivduals, where the class of the student is a covariate. The model
then simulates a SIR-epidemic on that network.

@@

The output of this model is \code{epimodel_sir}, an object of class
\code{netsim}, which contains a plethora of modeling information. We
provide the function \code{fortify_aggregate()}, which can take objects
from specialized classes of modeling output and transform it into a
tidy-style data frame.

@@

\hypertarget{a.-appendix}{%
\section*{A. Appendix}\label{a.-appendix}}
\addcontentsline{toc}{section}{A. Appendix}

\hypertarget{a.1-proof-of-theorem}{%
\subsection*{\texorpdfstring{A.1 Proof of Theorem
\ref{thm:sir-scale}}{A.1 Proof of Theorem }}\label{a.1-proof-of-theorem}}
\addcontentsline{toc}{subsection}{A.1 Proof of Theorem
\ref{thm:sir-scale}}

\begin{proof}\label{proof:thm}
\cite{Harko2014} provide an analytical solution for the Kermack and McKendrick equations (Eq. \eqref{eq:sir-ode}) by reparameterizing the ODEs so that $\mathcal{S}(u) = S(t)$, $\mathcal{I}(u) = S(t)$, and $\mathcal{R}(u) = R(t)$ for $0< u_T < 1$ with
\begin{align}\label{eq:harko-odes}
\mathcal{S}(u) &= S(0)u\\
\mathcal{I}(u) &= N - R(0) + NR_0^{-1}\log u - S(0)u \nonumber\\
\mathcal{R}(u) &= R(0) - NR_0^{-1} \log u, \nonumber
\end{align}
and $u$ and t are related by the following integral,
\begin{align*}
    t &= \int_{u}^1 \frac{N}{\beta \tau (N - R(0) + R_{0}^{-1} \log \tau - S(0)\tau)}d\tau \\
    &= \int_{u}^1 \frac{1}{\beta f(S(0), R(0), N, R_0, \tau)} d \tau\\
    &= \int_{u}^1 \frac{1}{\beta f(\tau)} d\tau,
\end{align*}
where we have made the denominator of the integral a function of $N$, the initial values, $R_0$, and $\tau$, which we further condense to $f(\tau)$ for brevity.
Then for a given $t$ we want to find $s$ such that $(S_1(t), I_1(t), R_1(t)) = (S_2(s), I_2(s), R_2(s))$.  Or equivalently, for a fixed $u$ want to find $v$ such that  $\mathcal{S}_1(u) = \mathcal{S}_2(v)$ and then the corresponding $t$ and $s$ are given by
\begin{align*}
    t & = \int_{u}^1 \frac{1}{\beta_1 f(\tau)} d\tau \\
    s & = \int_{v}^1 \frac{1}{\beta_2 f(\tau)} d\tau.
\end{align*}
Note that since the equations in Eq. \eqref{eq:harko-odes} are functions of the initial values and $R_0$, then $u = v$. We then can find a relation for $s$,
    \begin{align*}
    s & = \int_{u}^1 \frac{1}{\beta_2 f(\tau)} d\tau  \\
    & = \int_{u}^1 \frac{1}{a\beta_1 f(\tau)} d\tau \\ 
    &= \frac{1}{a}\int_{u}^1 \frac{1}{\beta_1 f(\tau)} d\tau \\
    &= \frac{1}{a}t.
\end{align*}
\end{proof}
